---
title: "Introduction to modLWFB90"
output: 
  rmarkdown::html_vignette:
    number_sections: true
    toc: true

vignette: >
  %\VignetteIndexEntry{Introduction to modLWFB90}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shall serve as an introduction to the *modLWFB90* package. It shows how to use its essential functions and how to specify certain aspects if desired.

  
  
  
***
# Introduction

The hydrological model LWFBrook90 is increasingly used at the FVA-BW in various projects and applications. It is a dynamic 1-D model that, through a large number of settings, allows for a detailed representation of soil, climate, and vegetation conditions. Hence, it is pretty useful for soil water modeling in the forest background we're working in. 

Paul Schmidt-Walter at the NW-FVA developed an R-package that serves as an R-wrapper for LWFBrook90.exe. *LWFBrook90R* makes it possible and relatively easy to use and work with LWFBrook90 from within Rstudio.

The package in front of you, *modLWFB90*, even goes one step further: It uses the whole functionality of *LWFBrook90R*, but adds a link to soil data and climate data available on the FVA-BW servers. This way users can do hydrological modelling by just giving the desired methods as arguments. All you need is one tidy data frame with the points you want to model in the following form:

* `ID_custom`   - a unique ID that all intermediate products as well as the results will be assigned to 
* `easting` and 
* `northing`    - coordinates in UTM 32N EPSG:32632
 
Then, all you need to specify is the time period, a number of settings (the origin of the soil data, the PTF to be used, the type of forest you want to model, etc.), and the way you want the output to be structured.
  
  
  
***
# Available Settings
 
## Vegetation
In the course of the project WHH-KW, parametrisation settings for the main tree species (beech - Buche, spruce - Fichte, oak - Eiche, pine - Kiefer) were agreed upon and can be selected in modLWFB90.

## Climate data
For WHH-KW the UHH Hamburg created a new and improved climate data set containing regionalised 250m-raster climate data for the historic time period 1961-2018, as well as the following climate scenarios: 

## Soil data
The following soil options can be selected: 

* STOK - Data from the Standortskartierung (add description...)
* BZE  - regionlised BZE-II data (add description...)
* OWN  - you can provide your own soil data

There's also a function to compare the three soil data outputs.    
  
  
  
***
# Basic Usage
 
## Getting the packages up and running
First of all, get the latest release of LWFBrook90R (at the time of creation v.0.4.0 from github). [github:pschmidtwalter/LWFBrook90R](https://github.com/pschmidtwalter/LWFBrook90R) 
Then, install modLWFB90, together with the dependency packages, which are required and will be installed automatically. The easiest way to install from github is with the following line of code: 

```{r, eval = F}
if(!require("remotes")) install.packages("remotes")
# HTTPS not supported when installing from github:  
Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS=TRUE)
Sys.setenv(HTTPS_PROXY=Sys.getenv("HTTP_PROXY")) # hier deine IP einfuegen

remotes::install_github(repo="pschmidtwalter/LWFBrook90R") 
remotes::install_github(repo="rhabel/modLWFB90")
```


```{r, warning = F, message = F}
library(modLWFB90)
```

As https is not supported from FVA-servers, this first has to be overwritten 
You might encounter issues with updates that can't be overwritten and other R-package-problems, in which case the following function might help you install the missing packages quicker (no install.packages(x); library(x) over and over again): 

```{r, warning = F, message = F}
check_packages <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)

}

# example: 
check_packages(c("LWFBrook90R", "parallel"))
```
  
  
## Input opjects
Essential for hydrological modelling are three datasets: Soil, Climate, and Parameters. This is how to get them the simplest possible way. First, we need sample data. `test.ids.bds` includes five points in the forest near Lake Constance (Bodensee). 
```{r}
test.ids.bds
```

  
### Soil data
The function to create soil data is called fnc_get_soil and returns a list of soil data frames, so best assign it to a list object like this: 
```{r, warning = F, message = F}
ls.soil <- fnc_get_soil(df.ids = test.ids.bds, 
                        soil_option = "STOK", 
                        PTF_to_use = "WESSOLEK", 
                        limit_MvG = T, 
                        limit_bodtief = T)
ls.soil[[1]]
```

There's a number of options to choose from: 

* df.ids - your data frame with `ID_custom`, `easting` and `northing` 
* soil_option - either `STOK`, `BZE`, (`STOK_BZE` for or `OWN`, if own you need to specify own soil data through df.soils. 
* PTF_to_use - either `HYPRES`, `PTFPUH2` or `WESSOLEK`
* for further settings (e.g. max soil depth or roots distribution) see `?fnc_get_soil`

***
There's also the option to compare the soil data created by the different options. Plus, the option to compare the hydraulic parameters created by the differnt PTFs. See `?fnc_compare_soil` for more information.
  
  
  
***
### Climate data
The function to create climate data is called `fnc_get_climate` and works the same way: 
```{r }
ls.clim <- fnc_get_clim(df.ids = test.ids.bds, 
                        mindate = as.Date("2010-01-01"), 
                        maxdate = as.Date("2011-12-31"))
```
  
Here you can only specify the time period for which you need climate data. The additional arguments `path_std` and `path_climdb` lead to the current position of the climate databases on the FVA-Servers. They may have to be adjusted at some point, but for now they are up to date and shouldn't be changed.   
   
If you model many points, large climate data files might use up all your RAM. In this case you should have a look at `fnc_read_climdb` and `fnc_write_climdb`. 

  
  
  
***
### Parameters
The list of parameters you can pass to LWFBrook90 is pretty exhaustive. See `LWFBrook90R::setparam_LWFB90` for a complete list. In the project WHH-KW, a set of parameters was agreed upon for the most relevant tree species (Buche - `beech`, Eiche - `oak`, Fichte - `spruce`, Kiefer - `pine`). They can be used by specifying the argument `tree_species` either as one value for all points, or as a vector of values for each individual point. Additional information on the modelling points can be passed with the argument `df.ind.info`. See `?fnc_get_params` to find out how to do this.  
  
  The default setting for spruce would look like this: 

```{r, warning=F}
ls.parms <- fnc_get_params(df.ids = test.ids.bds, 
                           tree_species = "spruce")
```


  
  
  
***
### Options
Modelling options are set with the function `set_optionsLWFB90` from the *LWFBrook90R* package. If you use *modLWFB90* the only settings you need to adjust are

* `budburst.method` and `leaffall.method` - if you want something other than `fixed`
* `startdate` and `enddate` - as desired, but should match your settings in `fnc_get_clim`
* `root.method` 

check `?set_optionsLWFB90` for more information.


***
### Adjust Output
Finally, we need to define how our output should look like. LWFBrook90 creates a lot of output files. In order to keep data storage to a minimum, both `run_LWFB90` and `run_multisite_LWFB90` provide an `output_fun` - argument that can be used to reduce the output and directly write it. 
   This is what the functions `fnc_write` (daily values) and `fnc_write_agg` (options to aggregate over year, vegetation period, or months) function are made for. They write temporary files with the desired output for each point. `fnc_write_to_sql` can be used to convert these files into a SQLite-DB. This "step-in-between" is necessary because SQLite does not support parallel writing. If you model few enough points that your memory can handle them, you can set dir_name to `NA` and get the results loaded into your environment. See the help pages of the write-functions on how to do this.

  
***
# Example
```{r, eval = FALSE}
# create ID-dataframe
df.ids = test.ids.bds

# set options
options.b90 <- set_optionsLWFB90(budburst.method = "Menzel",
                                 leaffall.method = "vonWilpert",
                                 startdate = as.Date("2010-01-01"),
                                 enddate = as.Date("2011-12-31"),
                                 root.method = "betamodel")
# set df.output
df.output <- set_outputLWFB90()
df.output[,] <- 0L
df.output[c("Evap", "Swat", "Flow"), c("Day")] <- 1
df.output[c("Flow"), c("Mon")] <- 1

output_path <- paste0(getwd(), "/")


# create data
ls.parms <- fnc_get_params(df.ids = test.ids.bds,
                           tree_species = "spruce")
ls.clim <- fnc_get_clim(df.ids = test.ids.bds)
ls.soil <- fnc_get_soil(df.ids = test.ids.bds,
                           soil_option = "STOK",
                           testgebiet = "BDS",
                           PTF_to_use = "WESSOLEK")

# this chunk of code removes non-mapped STOK-points, if there are any and 
# the BZE-filling option was not selected. Otherwise it doesn't do anything.
ls.clim <- ls.clim[which(!unlist(lapply(ls.soil, is.null))==T)]
ls.parms <- ls.parms[which(!unlist(lapply(ls.soil, is.null))==T)]
ls.soil <- ls.soil[which(!unlist(lapply(ls.soil, is.null))==T)]


# Run: 
  res <- run_multisite_LWFB90(options_b90 = options.b90,
                              param_b90 = ls.parms,
                              soil = ls.soil,

                              climate = ls.clim,

                              output = df.output,

                              all_combinations = F,
                              cores = parallel::detectCores(),
                              # if you have more than one core available, this will speed things up significantly

                              output_fun = fnc_write_agg,
                              aggr_tp = "vegper",
                              dir_name=paste0(output_path, "tmp/"))

  # save soils 
  names(ls.soil) <- NULL
  soils <- bind_rows(ls.soil)

  # create dirs
  if(!dir.exists(paste0(output_path, "tmp/soils/"))){
    dir.create(paste0(output_path, "tmp/soils/"), recursive = T)
  }
  save(soils, file = paste0(output_path, "tmp/soils/soils.RData"))

  # write to db --------------- ####
  fnc_write_to_sql(db_name = paste0(output_path, "test_db.sqlite"),
                   dir_tmp = paste0(output_path, "tmp"))

```
